---
output: github_document
bibliography: vignettes/articles/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MortIndEst

This package includes several functions for indirect estimates, specially about adult mortality.

Main development was done during World Population Prospects revision 2024, with the goal to include as inputs in the B3 model [@wpp2024].

Functions included are related to intercensal survival estimates, orphanhood, siblings survival, widowhood survival, census method for ${}_{15}q_{60}$ and old ages extension of mortality rates. Some of the contents consists in R implementations of materials in @iussp2012 and @un_manualX.

The idea of the package is to continue expanding its content, so suggestions are welcome. For other indirect methods not included here you can see [fertsestr](https://github.com/josehcms/fertestr), ...

## Installation

You can install the development version of MortIndEst from [GitHub](https://github.com/) with:

```{r, eval=F}
# install.packages("pak")
pak::pak("IvanWilli/MortIndEst")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, eval=F}
library(MortIndEst)
```

## References
