---
output: github_document
bibliography: vignettes/articles/references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# MortIndEst

This package includes several functions for indirect estimates of mortality, specially about adult mortality. Main development was done during World Population Prospects 2024 revision [@wpp2024].

Functions included are related to intercensal survival estimates, orphanhood, siblings survival, widowhood survival, child mortality, extinct generation methods and old ages extension of mortality rates. Some of the contents consists in R implementations of materials from @iussp2012 and @un_manualX.

See the [web](https://ivanwilli.github.io/MortIndEst/index.html) for applications, or see the examples in the function´s documentation.

## Installation

You can install the development version of MortIndEst with:

```{r, eval=F}
# install.packages("remotes")
remotes::install_github("IvanWilli/MortIndEst")
```

## References
