---
output: github_document
bibliography: vignettes/articles/references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# MortIndEst

This package includes several functions for indirect estimates, specially about adult mortality. Main development was done during World Population Prospects revision 2024, with the goal to include as inputs in the B3 model [@wpp2024].

Functions included are related to intercensal survival estimates, orphanhood, siblings survival, widowhood survival, census method for ${}_{15}q_{60}$ and old ages extension of mortality rates. Some of the contents consists in R implementations of materials from @iussp2012 and @un_manualX.

See the [web][https://ivanwilli.github.io/MortIndEst/index.html] for an application to Brazil, or see the examples in the function´s documentation.

## Installation

You can install the development version of MortIndEst from [GitHub](https://github.com/) with:

```{r, eval=F}
# install.packages("remotes")
remotes::install_github("IvanWilli/MortIndEst")
```

## References
