---
title: "Intercensal survival method for adult mortality: an application to Brazil"
bibliography: references.bib
---

```{r, include = FALSE}
# add a comment
pkgload::load_all()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", warnings = FALSE, messages = FALSE
)
```

# Introduction

Intercensal survival methods are based on relating the size of cohorts between two consecutive census dates, drawing on the population balancing equation for closed populations [@Hill2021; @Preston2001]. The main source of information is the population distribution by sex and age group from two successive censuses, generally conducted 5 or 10 years apart [@Arriaga2011; @Hill2021]. These methods are essential for producing estimates of adult mortality in contexts where vital registration systems are nonexistent or have insufficient coverage [@Hill2021; @IUSSP2013]. They also serve as diagnostic tools for assessing the consistency of demographic data and detecting anomalies in the age structure of the population [@Hill2021; @IUSSP2013].

The intercensal relationship between cohorts, assuming that the first census was conducted at ($t_1$) and the second at ($t_2$), is:

$$
{}_5S_x(t_1,t_2) = \frac{{}_5N_{x+(t_2-t_1)}(t_2)}
{{}_5N_x(t_1)}
$$

The main methodological variants include:

* **Model-pattern fitting (*match*):** this method is based on selecting a model life table family and identifying the level that provides the best fit for each (${}_5S_x$) corresponding to each age group. The indicators of interest are then obtained from the resulting life table (in our case ${}_{45}q_{15}$) and a summary measure is calculated across all cohorts.

  This is considered the most rudimentary method and was originally recommended in *Manual X* [@UnitedNations1983]. A later refinement presented by Preston [-@Preston1983] applies the Brass logit transformation, using the selected model family as the standard and the resulting level obtained from the fitting method as the reference [@Preston2001].

* **Forward and backward projection (*fproj* & *bproj*):** these methods start from the first—or the last—census and use a model family to project the population forward—or backward—in order to determine the mortality level required to reproduce the age structure of the second—or first—census. The literature indicates that differences between the two approaches will always exist, regardless of the errors present. Nevertheless, “backward projections provide a less biased estimate of the correct mortality level, provided that there are no age-reporting errors and that enumeration completeness differs between the two censuses” [@PalloniKominski1984]. In general, it is recommended that the method be applied in both directions and that a way of combining the results be sought, taking into account the particular characteristics of the population under study [@UnitedNations1983].

* **Variable-r (*var-r*):** the population by age and sex can be expressed as a function of survival ratios and cohort growth rates, a relationship derived by relaxing the stable-population assumption of a single growth rate originating from births [@BennettHoriuchi1981]. Given the size of each cohort at two different points in time and the average annual growth rate by age, it is possible to infer the implicit survival ratio for each age group [@PrestonBennett1983]. The main advantage of this method is that the width of the age groups does not need to coincide with the length of the intercensal period.

* **Feeney method (*feeney*):** the central idea of Feeney’s method [@UnitedNations2002] is to convert survival ratios into probabilities “rooted” at age 2.5. The procedure has different versions depending on the length of the intercensal interval—5 years, 10 years, or any other number of years.

The main limitation of these methods is their high sensitivity to changes in census coverage completeness between one census and the next. If the second census is more complete than the first, mortality will appear artificially low [@Hill2021; @IUSSP2013]. In addition, because the methods assume a closed population, unrecorded international migration biases the estimated number of deaths [@Hill2021; @Preston2001]. These situations produce cohort ratios greater than one, which must be excluded as inputs to the methods, at least in those methods that cannot be modified to incorporate migration. Finally, age exaggeration among older adults and age misreporting generally produce inconsistent or physically impossible survival ratios, such as values greater than one [@Hill2021; @Preston2001].

For step-by-step descriptions of the procedures, see Chapter XIV of *Manual X* [@UnitedNations1983] or Chapter 9 of Preston [-@Preston2001].

# Input data

Data that we need (both loaded with the package):

- Census population by five age-groups and sex in consecutive census

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
loc_name <- "Brazil"
census_data <- all_censuses |>
  filter(LocName == loc_name, SexName %in% c("Female", "Male")) |>
  arrange(SexName, TimeMid, AgeStart) |>
  select(LocName, IndicatorName, StatisticalConceptName, DataProcessType,
         TimeLabel, TimeMid, SexName, AgeStart, AgeSpan, DataValue)
```

- Life table information about infant/child mortality and $e_0$ (some methods requiere additional information)

```{r}
head(lt5_wpp24_brazil, 2)
```

# Application

First we need to organize the pair-census comparison.

```{r}
census_pairs <- tibble(
  date1 = head(sort(unique(census_data$TimeMid)), -1),
  date2 = tail(sort(unique(census_data$TimeMid)), -1)) |>
  mutate(t_middle = trunc((date1 + date2) / 2))
```

The for each one we will apply the different methods included in the `intercensal_survival` function, with ages of fit for each one (deafult based in literature, but can be also an additional combinatoric variable) and all the eleven model life table families as pattern (see `DemoToolsData::modelLTx1`).

```{r}
# combination of possible methods and census pairs
grid <- crossing(
  census_pairs,
  sex_name = c("Female", "Male"),
  tibble(
    method = c("match", "bproj", "fproj", "var-r", "feeney", "logit"),
    li = c(15, 0, 0, 10, 5, 0),
    ui = c(60, 40, 40, 30, 30, 60)
  )
)
```

We loop over the combinations of sex/interval/method/mlt with `pmap` from the `map` family functions (if we set `mlt_family = NULL` then results for all the models are returned). If you want to see in detail the application to some census pair (not looping in an industrial scale as here) see the `intercensal_survival` documentation.

```{r}
library(purrr)
runs <- pmap(grid, \(date1, date2, t_middle, sex_name, method, li, ui) {
  tryCatch({
    # get census data for the combination
    census_pair <- census_data |>
      filter(SexName == sex_name, TimeMid %in% c(date1, date2)) |>
      arrange(TimeMid, AgeStart)
    # get life table data for the combination
    lt_input <- lt5_wpp24_brazil |>
      filter(Sex == sex_name, Time == t_middle, AgeGrpStart %in% c(0, 1, 5)) |>
      arrange(AgeGrpStart)
    # apply methods
    out <- intercensal_survival(
      c1 = census_pair$DataValue[census_pair$TimeMid == date1],
      c2 = census_pair$DataValue[census_pair$TimeMid == date2],
      date1 = date1,
      date2 = date2,
      age1 = census_pair$AgeStart[census_pair$TimeMid == date1],
      age2 = census_pair$AgeStart[census_pair$TimeMid == date2],
      sex = if_else(sex_name == "Female", "f", "m"),
      mlt_family = NULL,
      method = method,
      ages_fit = seq(li, ui, 5),
      mlt_e0_logit_feeney = lt_input$ex[lt_input$AgeGrpStart == 0],
      q01_q05 = 1 - lt_input$lx[lt_input$AgeGrpStart %in% c(1, 5)] / 1e5,
      verbose = FALSE
    )
    # return
    list(
      result = out$selected_level |>
        left_join(out$Adult, by = "type") |>
        mutate(loc_name, sex_name, method, date1, date2, t_middle, .before = 1),
      error = NA_character_
    )
  }, error = \(e) list(result = NULL, error = conditionMessage(e)))
})
```

# Results

First we can take a look to combinations that did not work, and if we see something we need to loof for the reason (can exists quality problems in census data, or some issue arrangin the data, etc.). In ths case we are safe.

```{r}
issues_df <- grid |>
  mutate(error = map_chr(runs, "error")) |>
  filter(!is.na(error))
head(issues_df, 2)
```

There are many ways to get summary information from the results. We choose here to get min/max and inter-quartil percentiles for each sex, method and census-pair.

```{r}
results_df <- map2_dfr(runs, seq_len(nrow(grid)), \(x, i) {
  if (is.null(x$result)) return(NULL)
  x$result |> mutate(run_id = i, .before = 1)
})
dispersion_df <- results_df |>
  summarise(
    q15_45_min = min(q15_45, na.rm = TRUE),
    q15_45_p25 = quantile(q15_45, 0.25, na.rm = TRUE),
    q15_45_med = median(q15_45, na.rm = TRUE),
    q15_45_p75 = quantile(q15_45, 0.75, na.rm = TRUE),
    q15_45_max = max(q15_45, na.rm = TRUE),
    .by = c(sex_name, method, t_middle)
  )
```

We plot the results agaist WPP24 results (which in some way already considering the estimates). The `match` method yields higher adult mortality in all periods and for both sexes, while  `fproj` looks always lower, and the estimates produced by the other four methods are fairly similar. If the estimates from WPP 2024 are incorporated into the analysis, the mortality level estimated for 2000–2010 appears to be too low. Besides that, intercensal estimates can inform about generally ok the tendency between 1950-2010. 
Comparison with other estimates, regions, or countries is therefore particularly useful for calibrating the final adjustment.

```{r}
wpp_q15_45_df <- lt5_wpp24_brazil |>
  summarise(q15_45_wpp = 1 - lx[AgeGrpStart == 60] / lx[AgeGrpStart == 15],
            .by = c(Sex, Time))
ggplot(dispersion_df, aes(t_middle, q15_45_med, color = method)) +
  geom_linerange(aes(ymin = q15_45_min, ymax = q15_45_max), alpha = 0.5, linewidth = 0.8) +
  geom_linerange(aes(ymin = q15_45_p25, ymax = q15_45_p75), linewidth = 1.5) +
  geom_vline(xintercept = unique(census_data$TimeMid), col = "grey90", linetype = 2) +
  geom_point(size = 1.8) +
  geom_line(
    data = wpp_q15_45_df |> mutate(t_middle = Time + .5, sex_name = Sex),
    aes(x = t_middle, y = q15_45_wpp),
    color = "black",
    linewidth = 0.9,
    linetype = "dashed") +
  facet_wrap(~ sex_name) +
  labs(x = "Date", y = "q(15,45)", title = "Brazil intercensal survival estimates", subtitle = "Gery dashed lines are census dates") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor.x = element_blank())
```

# References